Phylogenetics and the human microbiome.

### 2013-09-24 sbreview.tex
Tree-thinking pass.

\documentclass{amsart}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}
\usepackage[round]{natbib}
\newcommand{\forarxiv}[1]{#1}
\newcommand{\notforarxiv}[1]{}

@@

Taxonomic issues.
Lack of "obvious" ways of grouping beasties.
Lack of broad phenotypic characterization.
These issues motivate sequence- and evolution-based methodology.

@@

\section{Methods used to investigate the human microbiome}

Terminology: survey versus metagenome.
Definition of a single sample.

Many of these methods cross over with microbial ecology in general.
Exploring prokaryotic diversity in the genomic era \cite{hugenholtz2002review}.

Gnotobiotic mice.
Human intervention.

\subsection{Microbial community estimation using marker genes}
History-- bacteria?

Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
This gene was chosen because...
They developed the first "universal" 16S PCR primers.
The great accomplishment of this is learning that archaea, although morphologically similar to bacteria, form their own separate lineage.

@@

Noise and the rare biosphere?

@@

Metagenomic data is typically used to infer information about metabolic capacity (cite Humann) rather than make statements of an evolutionary nature.
One approach to using metagenomic data is simply to subset to genes that are found in a large proportion of micro-organisms.
This is the approach taken by \citep{von2007quantitative,stark2010mltreemap} and Jonathan Eisen \citep{wu2008amphora} (in preparation).
Because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another, and even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.

@@

@@

% \cite{hummelen2010deep} Deep sequencing of the vaginal microbiota of women with HIV
% \cite{ravel2011vaginal} Vaginal microbiome of reproductive-age women
% \cite{fredricks2005molecular} Molecular identification of bacteria associated with bacterial vaginosis

@@

Phylogenetics has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
The most stunning example of this is the discovery of the archaea by Woese.
In the absence of sequencing, microbiologist have to rely on microscopes and staining.
Lots of little things look alike.

@@

\subsection{Diversity estimates using phylogenetics}

Because 16s surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature.

@@

@@

cite kembel paper

@@

The level of similarity between samples or groups thereof are often called \emph{beta diversity}; it is the current standard to employ phylogenetic tree structure in this sort of comparison.

@@

UniFrac distance and variants.
\cite{LozuponeKnightUniFrac05}
\cite{LozuponeEaWeightedUnifrac07}

\cite{BikEaMicrobiotaStomach06}

\cite{evans2012phylogenetic}
The phylogenetic Kantorovich-Rubinstein metric for environmental sequence samples

\cite{kuczynski2010microbial}
Microbial community resemblance methods differ in their ability to detect biologically relevant patterns

\cite{jaccard1908nouvelles}
Nouvelles recherches sur la distribution florale

@@

\cite{PurdomAnalyzingDataGraphs08}

\cite{nemergut2011global}
?? Global patterns in the biogeography of bacterial taxa

>>>>

\documentclass{amsart}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}
\usepackage{palatino}
\usepackage[round]{natbib}
\newcommand{\forarxiv}[1]{#1}
\newcommand{\notforarxiv}[1]{}

@@

Taxonomic issues.
Lack of "obvious" ways of grouping beasties.
Lack of broad phenotypic characterization.
In the absence of sequencing, microbiologist have to rely on microscopes and staining.
Lots of little things look alike.
These issues motivate sequence- and evolution-based methodology.

@@

\section{Methods used to investigate the human microbiome via sequencing.}

Terminology: survey versus metagenome.
Many of these methods cross over with microbial ecology in general \citep{hugenholtz2002review} (cite this?).
I will not be talking about manipulations of these systems, such as gnotobiotic mice and human intervention.

\subsection{Microbial community estimation using marker genes}

Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
This gene was chosen because...
They developed the first "universal" 16S PCR primers.

@@

@@

Some use metagenomic data as an expanded marker gene library.
\cite{kembel2011phylogenetic}
AMPHORA, Phylosift.
One approach to using metagenomic data is simply to subset to genes that are found in a large proportion of micro-organisms.
This is the approach taken by \citep{von2007quantitative,stark2010mltreemap} and Jonathan Eisen \citep{wu2008amphora} (in preparation).
Because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another, and even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.

@@

If not used as an expanded marker gene panel then typically used to infer information about metabolic capacity (cite Humann) rather than make statements of an evolutionary nature.

@@

@@

Phylogenetics has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
The most stunning example of this is the discovery that archaea, although morphologically similar to bacteria, form their own separate lineage.

@@

\subsection{Diversity estimates using phylogenetics}
Because 16s surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature.

@@

On expanded marker tree \cite{kembel2011phylogenetic}.

@@

@@

The level of similarity between samples or groups thereof are often called \emph{beta diversity};
Traditional way: \cite{jaccard1908nouvelles} Nouvelles recherches sur la distribution florale

UniFrac distance and variants has been very successful.
\cite{LozuponeKnightUniFrac05}
\cite{LozuponeEaWeightedUnifrac07}
\cite{kuczynski2010microbial} Microbial community resemblance methods differ in their ability to detect biologically relevant patterns

It recovers gradients.
\cite{nemergut2011global}
?? Global patterns in the biogeography of bacterial taxa

This has precedents.
\cite{evans2012phylogenetic}
The phylogenetic Kantorovich-Rubinstein metric for environmental sequence samples

\cite{BikEaMicrobiotaStomach06}
\cite{PurdomAnalyzingDataGraphs08}

@@

@@
